---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo = TRUE}
library(readr)
myData <- read_csv("activity.zip")
```



## What is mean total number of steps taken per day?
```{r echo = TRUE, tidy = TRUE, message = FALSE}
library(dplyr)

```

**Summarising the mean and median by interval across the dataset.**
```{r raw-summary, echo = TRUE, tidy = TRUE}
total_steps <- myData %>% group_by(date) %>% summarise(sum_steps = sum(steps)) %>% ungroup()
total_steps %>% summarise(mean = mean(sum_steps, na.rm = TRUE), median = median(sum_steps, na.rm = TRUE))
```

**Create a histogram with raw data**
```{r raw-histogram, echo = TRUE, fig.align = "center"}
hist(total_steps$sum_steps, main = "Average total steps", xlab = "Steps", breaks = 20)
```





## What is the average daily activity pattern?
```{r echo = TRUE, tidy = TRUE}
interval_steps <- myData %>% group_by(interval) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))
interval_steps
```

``` {r steps-breakdown, echo = TRUE, fig.align = "center"}
plot(interval_steps$interval, interval_steps$mean_steps, type = "l", main = "Average steps breakdown", xlab = "Interval", ylab = "Average number of steps")
interval_steps[which.max(interval_steps$mean_steps),]
```
On average, across all days in the dataset, interval `r interval_steps[which.max(interval_steps$mean_steps),"interval"]` contains the maximum number of steps, i.e., `r max(interval_steps$mean_steps)`.


## Imputing missing values
**The total number of missing values in the dataset**
```{r echo = TRUE}
length(myData[is.na(myData)])
```

**Imputing missing values by filling in the mean for that 5-minute interval**
```{r filled-summary, echo = TRUE}
filled_data <- myData
for (i in interval_steps$interval) {
    filled_data[which(filled_data$interval == i & is.na(filled_data$steps)), "steps"] <- interval_steps[which(interval_steps$interval == i), "mean_steps"]
}
filled_total_steps <- filled_data %>% group_by(date) %>% summarise(sum_steps = sum(steps)) %>% ungroup()
filled_total_steps %>% summarise(mean_steps = mean(sum_steps), median_steps = median(sum_steps))
```
After imputing missing values with the mean for each 5-minute interval, there is no change to the mean and a slight (minor) change to the median of the dataset.

**Create a histogram with imputed data**
```{r filled-histogram, echo = TRUE, fig.align = "center"}
hist(filled_total_steps$sum_steps, main = "Average total steps (missing data imputed)", xlab = "Steps", breaks = 20)
```

Comparing the two histograms, the general shapes of the two are same. The only difference is the frequency of the *mean* value due to the method of imputation.


## Are there differences in activity patterns between weekdays and weekends?
**Categorizing dates into weekdays and weekends**
```{r echo = TRUE}
filled_data <- filled_data %>% mutate(category = factor(ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekends", "weekdays")))
head(filled_data)
```

**Create a panel plot**
```{r panel-plot, echo = TRUE, fig.align = "center", message = FALSE, tidy = TRUE}
library(ggplot2)
filled_data %>% group_by(interval, category) %>% summarise(mean_steps = mean(steps)) %>% ggplot(aes(x = interval, y = mean_steps)) + geom_line() + facet_wrap(facets = vars(category), nrow = 2) + theme_light() + ylab("Average number of steps") + ggtitle("Average number of steps by weekdays/weekends", subtitle = "Missing data imputed") +  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```
There's a general pattern that during weekdays, a lot of steps are taken between 8-10 AM whereas the steps are more evenly spread during the weekends.

